{RILS}-{ROLS} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-12-01 paper_resources/doc/bmc_jbd/bmc_jbd.tex
Polishing up to statistical evaluation.

We propose a new meta-heuristic-based approach, called \textsc{RILS}-\textsc{ROLS} to solve this problem. \textsc{RILS}-\textsc{ROLS} is based on the following elements: ($i$) iterated local search is the method backbone, mainly solving combinatorial and some continuous aspects of the problem; ($ii$) ordinary least square method is focused on the continuous aspect of the search space -- it efficiently determines the best--fitting coefficients of linear combinations within solution equations;  ($iii$) a novel fitness function combines three important model quality measures: $R^2$ score, RMSE score, and the size of the model (or model complexity).

The experiments are conducted on the two well-known ground-truth benchmark sets called \textsc{Feynman} and \textsc{Strogatz}. \textsc{RILS}-\textsc{ROLS} was compared to 14 other competitors from the literature with  presence of different levels of the Gaussian white noise in the input data. Our method outperformed all 14 competitors with respect to the true symbolic model accuracy percentage under all level of noise. 
In addition to evaluation on known ground-truth datasets, a new randomly generated set of problem instances is introduced. The goal of \textsc{Random} dataset was to test the scalability of our method with respect to incremental equation sizes and different levels of noise. \textsc{RILS}-\textsc{ROLS} can be considered as a new state-of-the-art method for solving the problem of symbolic regression when ground-truth equations are known.
\end{abstract}

@@

\subsection{Parameter tuning}
The \textsc{RILS}-\textsc{ROLS} algorithm employs two control parameters, $penalty_{size}$ and $tolerance_{error}$. The first parameter is empirically tuned for considered benchmarks to 0.001 -- this is also used as a default value in the corresponding Python package (package is described in \ref{sec:appendix-1}). The second parameter is negatively correlated with the expected level of noise in input data, i.e., for higher noise the $tolerance_{error}$ takes lower values and vice-versa.

\fxnote{ drugi parametar manje vise zavistan od ocekivanog 
	nivoa suma -- ne znam kako obrazloziti njegov tuning (mozda prema greski na trening skupu pa izabrati onaj za koji su R2 i RMSE bolji...). }

\subsection{Comparison with other methods on the ground-truth benchmark sets}
In this section, we evaluate our \textsc{RILS-ROLS} algorithm and compare it to the 13 other competitors from the literature. The numerical results are reported in terms of the three Tables~\ref{tab:comp_noise0}--\ref{tab:comp_noise001}. The results for the two ground-truth benchmark sets, \textsc{Feynman} and \textsc{Storgatz}, are given w.r.t.\   the following level of noise: 0.0 (no-noisy data), 0.001 (low-level noise), and 0.01 (high-level noise), respectively. In details, the white Gaussian noise is added to the target ($y$-axis) values as a fraction of the signal RMS value, applied also in~\cite{la2021contemporary}, i.e. for target noise level $\alpha$ we have
$$ y_{noise} = y + \epsilon, \epsilon \sim \mathcal{N}\left(0, \alpha \sqrt{\frac{1}{n} \sum _{i=1}^n{y_i^2}}\right),$$
where $n$ relates to the number of items in the input data.

Each table consists of three blocks. The performances of all (14) algorithms are displayed in respective rows.  
The first block gives the label of the algorithm  whose performance is reported. The second block reports by two columns the relative percentages of all (10) algorithm runs on all problem instances for which the exact model has been found by respective algorithm  on \textsc{Feynman} and \textsc{Strogatz} benchmark set, respectively. The third column of this block provides the overall   relative exact percentage of success of the algorithm over all runs on all problem instances from the both benchmark sets. The third block displays the results of the respective algorithm in terms of   final $R^2$ score which indicate  on accuracy of found solution. The solution is considered accurate if it     produced an $R^2$ score larger that  0.999.  Note that the termination of our algorithm may not always be triggered by reaching an appropriate $R^2$ score, but also due to the reached time limit. These kind of results are of a particular interest as they may point out on the presence of strong over-fitting of algorithms in contrast to the primary goal, that is finding the right model. This block consists of three columns that provide the relative percentage of algorithm runs on the instances where he algorithm is able to reach the desired accuracy (fulfilling the condition $R^2 > 0.999$ upon termination) on the both ground--truth benchmark sets separately, and the overall percentage ov all (10) runs on all considered problem instances.

In short, the exact results, thus the number of algorithm runs over all instances (in percentages) for which the output of respective algorithm is able to find the exact model, are displayed in the second block of each  table. The numbers of runs   (in percentages) for which an algorithm  delivers a precise enough model (i.e. the outcome satisfying $R^2>  0.999$) are displayed in the third block of each of the tables.

@@

\begin{itemize}
	\item Concerning the exact percentages in case when no noise is utilized in the input data, the best performing algorithm is our \textsc{RILS-ROLS}, able to find the solution that match to the right model on 57.84\% runs over all problems instances of the benchmark set \textsc{Feynman}; it was even more impressive in solving problem instances from benchmark set \textsc{Strogatz}, being successful on 83.57\% runs. The second best performing  approach is \textsc{AI-Feynman}, successful in 55.78\%   runs over all problem instances from \textsc{Feynman} set, and just 27.14\% runs over the problem instances from benchmark set   \textsc{Strogatz}, respectively. All other approaches are performing much worse, and none of them is able to delver the exact solution for more than 30\% runs over the instances of any of the two considered ground-truth benchmark sets. 
	\item   Concerning the exact results in presence of the   noise  level of 0.001, our \textsc{RILS-ROLS} is still performing very well, having found an exact model for 42.08\% runs considering all problem instances from the both ground-truth benchmark sets. The second best approach is \textsc{AI-Feynman} which is successful on 31.89\% runs over all problem instances; as  already noticed, this algorithm performs better when applied on the instances from \textsc{Feynman} benchmark set than it is the case of \textsc{Strogatz} benchmark set. Approach  \textsc{AFP-FE} shows to be robust w.r.t. noise so its performances only slightly deteriorate in comparison to the no-noise scenario. This puts it on the third place with overall accuracy of 21.23\%. Other approaches reach   accuracy bellow 20\%. 
	\item  Concerning the exact results when the presence of noise is large with  the noise level of 0.01,  our \textsc{RILS-ROLS} is still performing robustly, having found an exact model for  34.77\% runes over all problem instances considering the both ground-truth benchmark sets. The second best is   \textsc{AFP-FE} reaching 20\% correct solutions. \textsc{AI-Feynman} performances are degraded significantly under this level of noise, so it is ranked only 6th overall, reaching the accuracy of 12.61\%.

	\item When comparing performances of the algorithms in terms of algorithm runs for which the obtained final model fulfills an accuracy of $R^2 > 0.999$, in case when no noise is integrated, our \textsc{RILS}-\textsc{ROLS} is successful with the rate of success of 83.38\% (third place). However, in contrast with the exact model percentages comparisons from above, there are a few methods reaching   better percentages of success than our algorithm: \textsc{MRGP}, and \textsc{Operon}, with 92.69\%, and 86.92\%, respectively. On the other hand, the exact model percentages of these two approaches are rather low (0\% and 16\%, respectively). These two algorithms are therefore more appropriate in scenarios when the underlying (ground-truth) model is unknown, so-called black-box regression.

	\item  In the presence of the noise level of 0.001, our \textsc{RILS-ROLS} method has 78.62\% of a success rate (ranked on the third place).  As before, \textsc{MRGP} and \textsc{Operon} are showing good performances.

@@

	\item Following the above conclusions, our \textsc{RILS}-\textsc{ROLS} algorithm is more robust than the other competitors, which is confirmed by the produced results w.r.t.\ different levels of noise in the input data. Moreover, it can be considered as new state-of-the-art algorithm when one solves the problems of SR with known ground-truth models.

@@

		\includegraphics[  width=190pt,height=150pt]{plots/time-avg-noise0.001.eps}
		\caption{Level of noise equal to 0.001}
		\label{fig:noise0.001-time}
	\end{subfigure}
	\centering
	\begin{subfigure}[b]{0.47\textwidth}
		\includegraphics[  width=190pt,height=150pt]{plots/time-avg-noise0.01.eps}
		\caption{Level of noise equal to 0.01.}
		\label{fig:noise0.01-time}
	\end{subfigure}
	\caption{Average runtimes on the instance problems where the exact solution has  been reached (considered only those algorithms whose the exact percentage is at least 5\% of all runs).}
	\label{fig:time-avg-exact-noise}
\end{figure}

In Figure~\ref{fig:time-avg-exact-noise}, average runtimes of the algorithms on those instance problems from the both ground-truth benchmark sets for which respective exact model has  been found, are compared. The algorithms which obtained an exact percentage of at least 5\% (considering all 1300 runs) are considered as relevant  and therefore included here, while others are just omitted.  The runtimes of each algorithm  are averaged w.r.t\  different levels of noise, and presented by a bar plot. Algorithms are labeled on $x$-axis, while respective average runtimes (oer all runs) are shown on $y$-axis. Note that $y$-axis is logarithmically scaled. The following conclusions are drawn from here.

\begin{itemize}
	\item When considering the instance problems with no-noisy input data, the lowest average runtime  is reported for \textsc{RILS}-\textsc{ROLS} algorithm (228$s$). A slightly worse runtime is obtained by \textsc{Dsr} algorithm (232$s$).  However, note that the relative exact percentage produced by \textsc{RILS}--\textsc{ROLS} is far better from \textsc{Dsr} (60.62\% vs. 19.71\%). 	All remaining approaches deliver an order of magnitude larger average runtimes than that of the \textsc{RILS}--\textsc{ROLS} method in case of runs when the exact model has been reached.   
	\item  When considering the instance problems with the noise level of 0.001,    the lowest average runtime  to find the exact model is again delivered by \textsc{RILS}-\textsc{ROLS} algorithm (108$s$). The second best average runtime is produced by \textsc{Dsr}, but this time significantly larger (622$s$) from the former algorithm. 
	\item When considering the instance problems with the noise level of 0.01, a similar conclusion may be drawn as that from above. 
\end{itemize}

@@

\section*{Appendix}
\section{Overview of \textsc{RILS}-\textsc{ROLS} python package}\label{sec:appendix-1}

>>>>

We propose an efficient meta-heuristic-based approach, called \textsc{RILS}-\textsc{ROLS} to solve this problem. \textsc{RILS}-\textsc{ROLS} is based on the following elements: ($i$) iterated local search is the method backbone, mainly solving combinatorial and some continuous aspects of the problem; ($ii$) ordinary least square method is focused on the continuous aspect of the search space -- it efficiently determines the best--fitting coefficients of linear combinations within solution equations;  ($iii$) a novel fitness function combines three important model quality measures: $R^2$ score, RMSE score, and the size of the model (or model complexity).

The experiments are conducted on the two well-known ground-truth benchmark sets called \textsc{Feynman} and \textsc{Strogatz}. \textsc{RILS}-\textsc{ROLS} was compared to 14 other competitors from the literature. Our method outperformed all 14 competitors with respect to the true symbolic model accuracy percentage under varying levels of noise. It also showed to be the most efficient considering the average running time of reaching the exact model.  
In addition to evaluation on known ground-truth datasets, a new randomly generated set of problem instances is introduced. The goal of \textsc{Random} dataset was to test the scalability of our method with respect to incremental equation sizes and different levels of noise. \textsc{RILS}-\textsc{ROLS} can be considered as a new state-of-the-art method for solving the problem of symbolic regression when ground-truth equations are known.
\end{abstract}

@@

\subsection{Parameter tuning}

The \textsc{RILS}-\textsc{ROLS} algorithm employs two control parameters, $penalty_{size}$ and $tolerance_{error}$. The first parameter is empirically tuned for considered benchmarks to 0.001 -- this is also used as a default value in the corresponding Python package (package is described in the Appendix). The second parameter is positively correlated with the expected level of noise in input data, i.e., for higher noise the $tolerance_{error}$ should take higher values and vice-versa. More precisely, in case there is no noise, we used the default setting $tolerance_{error}=10^{-16}$, while in noisy scenarios we set $tolerance_{error}$ to the corresponding noise level (0.001 or 0.01). 
For setting $penalty_{size}$ and $tolerance_{error}$ on new datasets one can either keep the default settings or use some wrapper tuning algorithm around \textsc{RILS-ROLS} regressor, e.g., grid-search w.r.t. training or validation accuracy.

\subsection{Comparison to other methods}
In this section, we evaluate our \textsc{RILS-ROLS} algorithm and compare it to 13 other competitors from the literature. The numerical results are reported in Tables~\ref{tab:comp_noise0}-\ref{tab:comp_noise001}. The results for \textsc{Feynman} and \textsc{Storgatz} are given w.r.t.\  the following levels of noise: 0.0 (no-noisy data), 0.001 (low-level noise), and 0.01 (high-level noise). More precisely, the white Gaussian noise is added to the target variable the same way as in~\cite{la2021contemporary}:
$$ y_{noise} = y + \epsilon, \epsilon \sim \mathcal{N}\left(0, \alpha \sqrt{\frac{1}{n} \sum _{i=1}^n{y_i^2}}\right),$$
where $n$ relates to the number of items in the input data.

Each table consists of two blocks, where the first one gives the exact model accuracy, while the second is related to the less relevant statistical accuracy based on the $R^2$ metric. The exact model accuracy shows how often the method reaches true formula (ground-truth). Regarding statistical accuracy, the solution is considered accurate if it produced an $R^2$ score larger than 0.999.  Note that for each instance there were 10 runs, so in total there were at most 1160 successful runs for \textsc{Feynman} and 140 for \textsc{Strogatz}.

@@

\begin{itemize}
	\item Concerning the exact percentages in case when no noise is utilized in the input data, the best performing algorithm is our \textsc{RILS-ROLS}. It found the exact solution in 57.84\% of runs for \textsc{Feynman} instances. It was even more impressive in solving \textsc{Strogatz} instances, being successful in 83.57\% runs. The second best performing  approach is \textsc{AI-Feynman}, successful in 55.78\% runs for \textsc{Feynman} set, and 27.14\% runs for the problem instances from benchmark set \textsc{Strogatz}. All other approaches perform significantly worse, and none of them is able to deliver the exact solution for more than 30\% runs over the instances of any of the two considered ground-truth benchmark sets. The overall results are in favor of \textsc{RILS-ROLS} with 60.62\%, while the second best is \textsc{AI-Feynman} with 52.65\%. 
	\item  Concerning the exact results in presence of the small noise (0.001), our \textsc{RILS-ROLS} is still performing very well, finding an exact model in 42.08\% of runs considering all problem instances. The second best approach is again \textsc{AI-Feynman} being successful in 31.89\% runs. As in no-noise scenario, there is a significant difference in \textsc{AI-Feynman} performances among two sets of instances -- it performs much better on \textsc{Feynman} instances than on \textsc{Strogatz}. It is also worth mentioning that \textsc{AFP-FE} shows to be robust w.r.t. noise so its performances only slightly deteriorate in comparison to the no-noise scenario. This puts it on the third place with overall accuracy of 21.23\%. Other approaches reach accuracy bellow 20\%.  
	\item  Concerning the scenario with high noise (level of 0.01), our \textsc{RILS-ROLS} is still performing quite well -- it finds and exact model in 34.77\% runs. The second best is \textsc{AFP-FE} reaching 20\% correct solutions. \textsc{AI-Feynman} performances degrade significantly under this level of noise, so it is ranked only 6th overall, reaching the accuracy of 12.61\%.

	\item One more interesting method is \textsc{DSR}. It shows the perfect robustness to noise and agnosticism to benchmark dataset -- it has near 19\% exact model accuracy in all 6 combinations of benchmark dataset (\textsc{Feynman} or \textsc{Strogatz}) and noise level (0, 0.001 or 0.01).

	\item When comparing performances of the algorithms in terms of $R^2 > 0.999$ accuracy, in no-noise scenario, our \textsc{RILS}-\textsc{ROLS} is successful with the rate of success of 83.38\% (third place). Better methods in terms of this accuracy are \textsc{MRGP}, and \textsc{Operon}, with 92.69\%, and 86.92\%, respectively. On the other hand, the exact model percentages of these two approaches are rather low (0\% and 16\%, respectively). These two algorithms are therefore more appropriate in scenarios when the underlying (ground-truth) model is unknown, so-called \emph{black-box} regression.

	\item  In the presence of the noise level of 0.001, our \textsc{RILS-ROLS} method has 78.62\% of a success rate (ranked third, again).  As before, \textsc{MRGP} and \textsc{Operon} are best two here.

@@

	\item We can conclude that \textsc{RILS}-\textsc{ROLS} achieves state-of-the-art performances regarding exact model accuracy. It also shows high robustness to noise. Regarding statistical accuracy, which is not so relevant for the ground-truth problems, \textsc{RILS}-\textsc{ROLS} also performs quite well.

@@

		\includegraphics[  width=190pt,height=150pt]{plots/time-avg-noise0.001.eps}
		\caption{Level of noise equal to 0.001}
		\label{fig:noise0.001-time}
	\end{subfigure}
	\centering
	\begin{subfigure}[b]{0.47\textwidth}
		\includegraphics[  width=190pt,height=150pt]{plots/time-avg-noise0.01.eps}
		\caption{Level of noise equal to 0.01.}
		\label{fig:noise0.01-time}
	\end{subfigure}
	\caption{Average runtime when exact solution is reached (considering only algorithms whose exact percentage is at least 5\% of all runs).}
	\label{fig:time-avg-exact-noise}
\end{figure}

Figure~\ref{fig:time-avg-exact-noise} compares the methods w.r.t. average runtimes (in seconds) when exact solution is reached. We do not consider methods whose exact percentage is too low, i.e., bellow 5\%. Methods are labeled on $x$-axis, while respective average runtimes are shown on $y$-axis. (Note that $y$-axis is logarithmically scaled.) The following conclusions are drawn from these results.

\begin{itemize}
	\item When considering the instance problems with no-noise, the lowest average runtime is reported for \textsc{RILS}-\textsc{ROLS} algorithm (228 sec.). A slightly worse runtime is obtained by \textsc{DSR} algorithm (232 sec.). Note that the relative exact percentage produced by \textsc{RILS}-\textsc{ROLS} is far better from \textsc{DSR} (60.62\% versu 19.71\%). All remaining approaches deliver an order of magnitude larger average runtimes than that of \textsc{RILS}-\textsc{ROLS} and \textsc{DSR} methods..   
	\item  When considering instances with the noise level of 0.001, the lowest average runtime to find the exact model is again delivered by \textsc{RILS}-\textsc{ROLS} algorithm (108 sec.). The second best is produced by \textsc{DSR}, but this time significantly larger (622 sec.) compared to \textsc{RILS-ROLS}. 
	\item When considering the instance problems with the noise level of 0.01, the conclusion is similar to previous. 
\end{itemize}

@@

\section*{Appendix -- \textsc{RILS}-\textsc{ROLS} Python package}\label{sec:appendix-1}
